Longitudinal Principal Manifold Estimation

### 2023-12-16 paper/jsm_2024.tex

\usepackage[
backend=biber,
style=authoryear
]{biblatex}

\addbibresource{references.bib}

@@

\begin{algorithm}
\caption{Longitudinal Principal Manifold Estimation}\label{alg:lpme}
  \KwData{Data points $\left\{Y_{i, t}\right\}_{i=1, t=1}^{I_t, T}$, positive integer $d$, positive integer $N_0 < \min{I_t} - 1$, $\alpha$, $\epsilon$, $\epsilon^* \in (0, 1)$, candidate tuning parameters $\left\{\lambda_k\right\}_{k=1}^K$, $\left\{\gamma_l\right\}_{l=1}^L$, $itr \geq 1$, which is the maximum number of iterations allowed.}
\KwResult{Analytic formula of $\hat{f}^*: \mathbb{R}^{d + 1} \to \mathbb{R}^{D + 1}$, optimal tuning parameter $\gamma^*$.}
\For{$t = 1, 2, \dots, T$} {
  Apply HDMDE algorithm with input $\left(\left\{Y_{i, t}\right\}_{i = 1}^{I_t}, N_0, \epsilon, \alpha\right)$ and obtain $N_t$, $\left\{\mu_{j, t}\right\}_{j = 1}^{N_t}$, and $\left\{\theta_{j, t}\right\}_{j = 1}^{N_t}$\;
}

@@

Notably, in many applications where manifold learning is implemented, the high-dimensional shapes of interest are self-intersecting manifolds. Methods based on the principal curve framework, including the principal manifold approach used here, are typically unable to fit such manifolds due to the self-consistency condition. To avoid this issue, we propose a data augmentation approach with polar or spherical coordinates, depending on the value of $D$. As a simple example, the unit circle can be considered a self-intersecting manifold with $d = 1$ when viewed in two dimensions. However, by adding a third dimension equal to the angle of each observation from the origin, the manifold can be considered as having $d = 1$ and $D = 3$, while avoiding any self-intersections. This enables the PME algorithm, and thus the LPME algorithm, to be fit under these circumstances. This process is illustrated graphically in Figure \ref{fig:unit_circle_augmentation}.

\begin{figure}[ht]
  \centering
  \subfloat[\centering Unaugmented Unit Circle]{{\includegraphics[height=9cm]{unit_circle_D2}}}%
  \hfill
  \subfloat[\centering Augmented Unit Circle]{{\includegraphics[height=9cm]{unit_circle_D3}}}
  \caption{Illustration of our proposed data augmentation approach using the Unit Circle. The left panel (a) shows the depiction of the circle with $d=1$ and $D=2$, while the right panel (b) shows how our proposed use of considering this same manifold in $D=3$ by using the spherical coordinates results in a manifold that is not self-intersecting.}
  \label{fig:unit_circle_augmentation}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{sim_case1}
  \includegraphics[width=\textwidth]{sim_case5}
  \caption{Simulation Cases 1 and 5.}
  \label{fig:sim_case1}
\end{figure}

@@

As discussed previously, meaningful between-image error in estimates of subcortical structures is introduced during the process of segmenting MRI images. This section demonstrates how the LPME algorithm may be used to mitigate the effects of such noise on further analysis using these estimates. To achieve this, we used MRI data collected through the ADNI study, a longitudinal observational study with the goal of identifying imaging biomarkers to assess the progression of Alzheimer's disease (AD).

@@

Results of fitting the PME and LPME algorithms on the surface of the left hippocampus of a single participant are shown in Figures \ref{fig:adni_result}. Visual inspection indicates that while there are slight differences between the shape shown in the data and the shape estimated by the LPME algorithm, the surface estimated by LPME appears to fit well to the data. There are two main sources of discrepancies between the data and the LPME estimate. First, at time points where the observed surface changes orientation, as is visible between the first two time points, for example, the LPME estimates maintain a consistent orientation. The second difference between the LPME estimates and the observed data can be seen at the sharper corners of the hippocampus, where the LPME estimates do not fully capture the severity of the curves seen in the observed data.

Figure \ref{fig:adni_cross_section} depicts cross sections of the observed hippocampus surface and the estimated surface obtained using LPME and PME at each individual time point. These cross sections illustrate the differences between the estimates reached by the PME and LPME algorithms, with the LPME-based estimate being generally less responsive to changes in the shape and orientation of the hippocampus between time points. This figure reiterates that both the PME and LPME algorithms are unable to capture the distinctive sharp curve in the structure at the upper-left corner of each of the plots. Additionally, we see that, along the upper-right edge of the surface, there is a spot where the estimated manifold intersects with itself. This is present for both the PME and LPME estimates, but appears to be more prevalent when using LPME.

@@

Considering the same individual shown in Figures \ref{fig:adni_result} and \ref{fig:adni_cross_section}, results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figure \ref{fig:adni_lthal_cross_section}. While the segmentation estimates of the thalamus do not show the same sort of deviations that are clearly visible for the hippocampus, Figure \ref{fig:adni_lthal_cross_section} indicates that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.

@@

  \caption{Left Hippocampus, Cognitive Healthy ADNI Participant}
  \label{fig:adni_result}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \subfloat[\centering Left Hippocampus Cross Section]{\includegraphics[height=10cm]{results/adni_plots/adni_cross_section}}%
  \hfill
  \subfloat[\centering Estimated Volumes]{\includegraphics[height=10cm]{}}
  \includegraphics[height=10cm]{adni_plots/adni_cross_section}
  \caption{Left Hippocampus Cross Section}
  \label{fig:adni_cross_section}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[height=10cm]{adni_plots/adni_lthal_cross_section.png}
    \caption{Left Thalamus Cross Section}
    \label{fig:adni_lthal_cross_section}
\end{figure}

\begin{figure}%
  \centering
  \subfloat[\centering Data]{{\includegraphics[height=22cm]{adni_plots/adni_data_plot}}}%
  \hfill
  \subfloat[\centering LPME Estimate]{{\includegraphics[height=22cm]{adni_plots/adni_lpme_plot}}}

  \caption{Left Hippocampus, Cognitive Healthy ADNI Participant}
  \label{fig:adni_result}
\end{figure}

@@

Another limitation stems from the performance of the LPME method on the hippocampus surfaces obtained from the ADNI dataset. It is clear that neither the PME nor the LPME estimates were able to fully capture the distinctive curved structure of the hippocampus. This shortcoming could potentially be addressed in several ways. First, as discussed above, the manifold learning method used in the initialization step of the PME algorithm, or the parameters of that method could be adjusted to better meet the needs of the specific manifold being estimated. Second, while a simple grid search is used to select the optimal value of the smoothing parameter in both the PME and LPME algorithms, it is possible that an improved search method would yield results that are more sensitive to the nuances of the structure being estimated. Finally, it may be possible to manipulate the manifold to be easier to estimate through adjustments to the data augmentation process, as described in Section \ref{s:LPME}.

Additional work will be necessary to more thoroughly explore the data augmentation approach proposed in this article and its capabilities. The simple calculations used here allowed the PME and LPME algorithms to estimate self-intersecting manifolds that would not previously have been estimable by these methods. However, while the use of polar and spherical coordinates addressed a problem posed by well-behaved circular structures, it would not necessarily apply easily to more complex structures, such as a figure 8 shape, or even rounded structures with large variations in the radius, which would result in several points sharing an angle from the origin. The development of a procedure to handle these considerations would greatly expand the potential use cases for this approach, thus allowing principal manifold-based methods to be used in a wider range of situations. While the approach described in this article is limited to the estimation of manifolds from observations taken from a single structure over time, the further extension of this method to manifold estimation for groups of similar structures would also be of interest.

>>>>

\usepackage[
backend=biber,
style=authoryear,
url=false,
isbn=false
]{biblatex}

\addbibresource{references.bib}
\AtEveryBibitem{\clearfield{month}}
\AtEveryBibitem{\clearfield{day}}

@@

\begin{algorithm}
\footnotesize
\caption{Longitudinal Principal Manifold Estimation}\label{alg:lpme}
  \KwData{Data points $\left\{Y_{i, t}\right\}_{i=1, t=1}^{I_t, T}$, positive integer $d$, positive integer $N_0 < \min{I_t} - 1$, $\alpha$, $\epsilon$, $\epsilon^* \in (0, 1)$, candidate tuning parameters $\left\{\lambda_k\right\}_{k=1}^K$, $\left\{\gamma_l\right\}_{l=1}^L$, $itr \geq 1$, which is the maximum number of iterations allowed.}
\KwResult{Analytic formula of $\hat{f}^*: \mathbb{R}^{d + 1} \to \mathbb{R}^{D + 1}$, optimal tuning parameter $\gamma^*$.}
\For{$t = 1, 2, \dots, T$} {
  Apply HDMDE algorithm with input $\left(\left\{Y_{i, t}\right\}_{i = 1}^{I_t}, N_0, \epsilon, \alpha\right)$ and obtain $N_t$, $\left\{\mu_{j, t}\right\}_{j = 1}^{N_t}$, and $\left\{\theta_{j, t}\right\}_{j = 1}^{N_t}$\;
}

@@

Notably, in many applications where manifold learning is implemented, the high-dimensional shapes of interest are self-intersecting manifolds. Methods based on the principal curve framework, including the principal manifold approach used here, are typically unable to fit such manifolds due to the self-consistency condition. To avoid this issue, we propose a data augmentation approach with polar or spherical coordinates, depending on the value of $D$. As a simple example, the unit circle can be considered a self-intersecting manifold with $d = 1$ when viewed in two dimensions. However, by adding a third dimension equal to the angle of each observation from the origin, the manifold can be considered as having $d = 1$ and $D = 3$, while avoiding any self-intersections. This enables the PME algorithm, and thus the LPME algorithm, to be fit under these circumstances.

@@

\begin{figure}[h]
  \centering
  \includegraphics[height=11cm]{sim_case1}
  \includegraphics[height=11cm]{sim_case5}
  \caption{Simulation cases 1 and 5. The function estimates found to be closest to the true manifold function, shown in red, can be considered the best fitting. In each case, the LPME estimate in blue is closest to the true manifold over all time points.}
  \label{fig:sim_case1}
\end{figure}

@@

As discussed previously, meaningful between-image error in estimates of subcortical structures is introduced during the process of segmenting MRI images. This section demonstrates how the LPME algorithm may be used to mitigate the effects of these variations on further analysis. To achieve this, we used MRI data collected through the ADNI study, a longitudinal observational study with the goal of identifying imaging biomarkers to assess the progression of Alzheimer's disease (AD).

@@

Results of fitting the PME and LPME algorithms on the surface of the left hippocampus of a single participant are shown in Figure \ref{fig:adni_result}. Visual inspection indicates that while there are slight differences between the shape shown in the data and the shape estimated by the LPME algorithm, the surface estimated by LPME appears to capture the primary patterns observed in the data. There are two main sources of discrepancies between the data and the LPME estimate. First, at time points where the observed surface changes orientation, as is visible between the first two time points, for example, the LPME estimates maintain a consistent orientation. The second difference between the LPME estimates and the observed data can be seen at the sharper corners of the hippocampus, where the LPME estimates do not fully capture the severity of the curves seen in the observed data.

Figure \ref{fig:lhipp_cross_sections} depicts cross sections of the observed hippocampus surface and the estimated surface obtained using LPME and PME at each individual time point. These cross sections illustrate the differences between the estimates reached by the PME and LPME algorithms, with the LPME-based estimate being generally less responsive to changes in the shape and orientation of the hippocampus between time points. This figure reiterates that both the PME and LPME algorithms are unable to capture the distinctive sharp curve in the structure at the upper-left corner of each of the plots. Notably, it appears that the boundaries of the LPME-estimated surface tend to fall inside the boundaries of the PME-estimated surface, particularly at locations with the highest level of curvature.

To understand the impacts of these discrepancies on the estimated volume of the regions, the volume of the hippocampus was approximated from the raw data and the estimates obtained by the PME and LPME algorithms at each time point. The volume estimation approach used here relies on counting the voxels contained within the boundary defined by the PME- or LPME-estimated embedding map, with each voxel having a known volume. Further details of this approach are beyond the scope of this article, but the function code is available at \href{https://github.com/rjzielinski/lpme-project}{\texttt{https://github.com/rjzielinski/lpme-project}}. The volume estimates can be found in Figure \ref{fig:lhipp_volumes}. Here, we see that volume estimates differ substantially depending on whether PME, LPME, or the raw data were used, with the LPME estimate resulting in the lowest volume computations.

@@

Considering the same individual shown in Figures \ref{fig:adni_result} and \ref{fig:adni_cross_section}, results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figure \ref{fig:adni_lthal_cross_section}. While the segmentation estimates of the thalamus do not show the same sort of deviations that are clearly visible for the hippocampus, Figure \ref{fig:lthal_cross_sections} indicates that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus. This is reflected in the smaller discrepancies between volume estimates, as seen in Figure \ref{fig:lthal_volumes}, though visible variations between time points remain.

@@

  \caption{Left Hippocampus, Cognitive Healthy ADNI Participant. The raw surface data, displayed in red, show slight changes in orientation that are absent in the LPME estimates, shown in blue. The estimates obtained by the PME and LPME algorithms appear unable to accurately capture the true shape of the structure at points with high levels of curvature.}
  \label{fig:adni_result}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \subfloat[\centering Left Hippocampus Cross Section]{\label{fig:lhipp_cross_sections} \includegraphics[width=8.5cm]{adni_plots/adni_cross_section}}%
  \hfill
  \subfloat[\centering Estimated Volumes]{\label{fig:lhipp_volumes} \includegraphics[width=8.5cm]{adni_plots/lhipp_volume}}
  \caption{Left Hippocampus Cross Section}
  \label{fig:adni_cross_section}
\end{figure}

\begin{figure}[h]
  \centering
  \subfloat[\centering Left Thalamus Cross Section]{\label{fig:lthal_cross_sections} \includegraphics[width=8.5cm]{adni_plots/adni_lthal_cross_section}}%
  \hfill
  \subfloat[\centering Estimated Volumes]{\label{fig:lthal_volumes} \includegraphics[width=8.5cm]{adni_plots/lthal_volume}}
  \caption{Left Thalamus Cross Section}
  \label{fig:adni_lthal_cross_section}
\end{figure}

@@

Another limitation stems from the performance of the LPME method on the hippocampus surfaces obtained from the ADNI dataset. It is clear that neither the PME nor the LPME estimates were able to fully capture the distinctive curved structure of the hippocampus. This shortcoming could potentially be addressed in several ways. First, as discussed above, the manifold learning method used in the initialization step of the PME algorithm, or the parameters of that method could be adjusted to better meet the needs of the specific manifold being estimated. Second, while a simple grid search is used to select the optimal value of the smoothing parameter in both the PME and LPME algorithms, it is possible that an improved search method would yield results that are more sensitive to the nuances of the structure being estimated. Finally, it may be possible to manipulate the manifold to be easier to estimate through adjustments to the data augmentation process, as described in Section \ref{s:LPME}. The volume estimates obtained from the outputs of the PME and LPME algorithms reflected the insufficient fit of these algorithms to the hippocampus data. However, the volume estimation process includes ad hoc approaches to address missingness in the data, and the high level of variability in the estimated volumes of the thalamus despite visual indications of improved fit to this data suggests that this approach may itself be contributing to this variability. Further testing of this approach will be needed before the LPME algorithm will be viable for approximating the volume of brain regions.

Additional work will also be necessary to more thoroughly explore the data augmentation approach proposed in this article and its capabilities. The simple calculations used here allowed the PME and LPME algorithms to estimate self-intersecting manifolds that would not previously have been estimable by these methods. However, while the use of polar and spherical coordinates addressed a problem posed by well-behaved circular structures, it would not necessarily apply easily to the hippocampus or even more complex structures, such as a figure 8 shape, or even rounded structures with large variations in the radius, which would result in several points sharing an angle from the origin. The development of a procedure to handle these considerations would greatly expand the potential use cases for this approach, thus allowing principal manifold-based methods to be used in a wider range of situations. While the approach described in this article is limited to the estimation of manifolds from observations taken from a single structure over time, the further extension of this method to manifold estimation for groups of similar structures would also be of interest.
